Monitoring and predictive mapping of floristic biodiversity along a climatic gradient in ENSO's terrestrial core region, NW Peru

### 2020-08-10 text/ecography/round_1/review/reply_to_reviewers.Rmd

---
output:
  bookdown::pdf_document2:
    keep_tex: no
    latex_engine: xelatex
    # float is needed for floating figures
    extra_dependencies: ["fontspec", "sectsty", "float"]
    toc: false
    fig_caption: yes
    number_sections: true
header-includes:
  # adjust figure captions
  - \usepackage{caption}
  - \captionsetup{font={footnotesize,bf,singlespacing}}

title: "Point-by-point reply to reviewers' comments"
bibliography: ../../round_2/transect_refs.bib
csl: ../../round_2/ecography.csl
# biblio-style: "natbib"
link-citations: true

---

```{r setup, include=FALSE}
library("knitr")
library("kableExtra")
library("tidyverse")
library("here")
# make sure that figures are allowed to fload (hold the assigned position and are not always put on top of a page)
knitr::opts_chunk$set(fig.pos = 'H')
```

# Reply to subject editor Bethany Bradley

Dear Dr Bradley,

thank you very much for handling our manuscript.
We are pleased to resubmit a revised version of our manuscript now entitled  *Monitoring and predictive mapping of floristic biodiversity along a climatic gradient in ENSO's terrestrial core region, NW Peru*.
We are also very thankful for the constructive comments of both reviewers which helped to improve substantially our manuscript.
Please find below a point-by-point reply to all concerns raised by the reviewers.
In accordance with your and the reviewers' suggestions, we have:

- formulated clear hypotheses and adjusted accordingly the framing of the manuscript.
- explained our methods in much more detail.
To make our study truly reproducible, we will publish our study upon acceptance as a research compendium.
This means that we will make our code and text publicly available via [github](https://github.com).
The data will be published on DRYAD or Zenodo.
And a docker container will make available the software used during the analyses thereby ensuring that our code will run on any operating system and regardless of software updates.
- embedded our study even better in the ongoing scientific debate while focussing on the key points of our study.
- highlighted more prominently the novelty of our study in the introduction and discussion section.
- written out El Niño/La Niña in full rather than abbreviating.

Therefore, we hope that our manuscript is now suitable for publication.

Yours sincerely,

Dr Jannes Muenchow

# Reply to reviewer #1

1. **The authors assessed the effect of water availability on species diversity across a gradient in the Northwest Peru in different years in which the effect of climatic events (ENSO) have influence on plant diversity. They used 50 (0.09ha) plots distributed in different water availability conditions along a transect and measured plant identity and plant coverage of vascular plants in each one. In addition, they used an irrigation-experiment to assess under controlled conditions the effects of water availability and their interaction with nutrient enrichment on plant diversity accumulation. This integrative approach which combines macroecological analyses and experimental tests of the potential drivers is very interesting and rare in ecological studies. However, the methods and results are not clear. Particularly the description of the plot data, the modeling approaches, and the display of results. Moreover, there is a mismatch in the response variable between the observational analyses and the experiments. Finally, in the introduction, the hypotheses do not have clear predictions, and the discussion focusses on topics that are not presented as the focus of the manuscript in the introduction. In addition, the text should be reviewed to improve the quality of the writing, there are some sentences that are not clear. In my opinion, this study has a lot of potential, but these issues need to be addressed.**

First of all, thank you very much for your detailed feedback which helped to improve substantially the manuscript.
In accordance with your suggestions, we have rewritten the methods and results sections.
Additionally, we have described in much more detail the plot data and the modeling approaches.
We also show that in fact there is no mismatch between the response variables (please refer to the reply to comment #12).
We are sorry that this was not clear from the beginning.
We also have rewritten the research questions as hypotheses with clear predictions.
Additionally, we absolutely agree with you that parts of the discussion in the first draft of the manuscript were out of scope.
Thanks for pointing this out, we have deleted or largely rewritten the corresponding sections and the Conclusions.
Finally, we have improved the writing.

2. **Line 48: The introduction should include a bit about the expectation of species diversity response to climate conditions. Also, the term “ecosystem diversity” should be clarified.**

We have rewritten our research questions as follows:

>>> Specifically, we investigated the:
>>>
(1) change in the different levels of biodiversity, namely alpha, beta, gamma and ecosystem diversity. 
The latter was represented by ordination scores and additionally visualized by means of predictive mapping.
Overall, we expected a boost in biodiversity as soon as water was no longer a limiting factor.
Regarding the spatial distribution of the floristic composition we expected a more pronounced development of vegetation assemblages along the climatic gradient with wetter conditions.
>>>
(2) effects of soil characteristics and topography in shaping floristic composition and vegetation cover (biomass production). 
We expected that the influence of topographic and edaphic variables might gain in influence in wetter years.
>>>
(3) We complemented the plot survey at the landscape scale with an irrigation-fertilization experiment, thereby considering the socio-economic perspective of the ENSO variability. 
Specifically, we quantified the effect of the water-nutrient interaction on biomass development under different rainfall scenarios.
We expected that the water input of the rainfall scenario representing a Super-Niño event in combination with the fertilizer would be most conducive to biomass and biodiversity development.

Additionally, we now specifically say that we represented ecosystem diversity by ordination scores.

3. **Line 50: The introduction is focused on floristic composition and the second question also includes information on biomass production. Here, there are many studies that showed the effect of ENSO and extreme climatic events in aboveground biomass production (e.g. Phillips et al. 2009). Based on these previous studies, the authors could include a bit more of information on the response of biomass production and their expectations for the study system.**

Regarding our expectations, please refer to the reply to the previous comment.
Thanks for pointing us to relevant literature.
We have added following sentence to the introduction:

>>>In any case, ENSO certainly drives a large amount of the interannual variability of woody plant growth across the tropics [@rifai_enso_2018] though the specific response to El Niño may vary from region to region [@phillips_changes_2009].

<!--
- [@phillips_changes_2009] -> mixed response of Amazon forest to EN induced droughts.
However, climate change affects Amazon moist forests, there is a risk that they might be converted from carbon sinks to sources.
- cite also Salazar et al. (accepted) since they also show that there is a risk that climate change might convert TDF from sink to sources.
- [@rifai_enso_2018]: ENSO Drives interannual variation of forest woody growth across the tropics 
Overall, there is high correlation (r = -0.75) between the annual anomaly of tropical forest woody growth and the annual mean of the El Niño 3.4 index, driven mainly by strong correlations with anomalies of soil water deficit, vapour pressure deficit and shortwave radiation.

Climate change has the potential to convert both tropical moist as well as dry forests from carbon sinks to sources.
-->

4. **Line 86: The description of vegetation sampling should be improved. It is not clear how the proportion of area occupied by each species is measured. Is it based on basal area o crown dimensions?**

We have clarified as follows:

>>>Plant cover refers to the relative projected area covered by a species [@damgaard_estimating_2014].

5. **Line 100: Are there more than one plot within the same NDVI pixel?**

There is not one pair of plots which are located in the same pixel (see Table \@ref(tab:ndvi)).

```{r ndvi, echo=FALSE, eval=TRUE, message=FALSE}
# attach packages
library("knitr")
library("dplyr")
library("raster")
library("reshape2")
# attach raster stack
ndvi = readRDS(here("images/00_ndvi_rasters.rds"))
# attach env_data
env_data = readRDS(here("images/11_env_data.rds"))
env_data$year = gsub(".*_", "", env_data$pnr)

# just keep unique plots (could have also used the id-column)
data = env_data[!duplicated(paste(env_data$lon, env_data$lat)),
                  c("lon", "lat")]
# extract values from NDVI rasters to plot locations
vals = raster::extract(ndvi[[c("ndvi_11", "ndvi_12", "ndvi_16", "ndvi_17")]], 
                       data[, c("lon", "lat")])
# count different NDVI values for each sampled year
tab = melt(vals) %>%
  group_by(Var2) %>%
  summarize(no_plots = n(), distinct_ndvi_values = n_distinct(value))
tab = rename(tab, year = Var2) %>%
  mutate(year = gsub("ndvi_", 20, year))
kable(tab, "latex", booktabs = TRUE, escape = TRUE,
      caption = "Showing that each plot has a distinct NDVI value.")
```

6. **Line 101: Worldclim data are static and describe the general pattern of variation, but I´m not sure if this database is appropriate for this study. Can you explore the inclusion on TRMM data? These data have a coarse spatial resolution but a fine temporal resolution and could improve your analysis.**

```{r gam-fig, echo=FALSE, message=FALSE, results = "asis", }
library("mgcv")
require("itsadug")

# attach response (DCA scores)
data = readRDS(here("images/15_model_input_data.rds"))
data$id = as.numeric(data$id)
data$year = as.factor(data$year)
fo = as.formula(DCA1 ~ s(ndvi, by = year, k = 3) + year)
gam_1 = gam(fo, data = data)
# gamtabs(summary(gam_1))
# plot_smooth(gam_1, view = "ndvi", plot_all = "year", ylim = c(-2, 3))
```

We appreciate the recommendation.
However, the spatial resolution of TRMM data is too low for our data, i.e., one pixel would cover several plots.
Please note also that we already have used the MODIS data for the vegetation period of each year, i.e., we have used MODIS data from 2011, 2012, 2016 and 2017. 
This also means that the NDVI for each year was computed as the mean for all NDVI MODIS images available for the vegetation period (March to May). 
We say so now more clearly in the text as follows:

>>> We downloaded the normalized difference vegetation index (NDVI) raster imagery from the Moderate Resolution Imaging Spectrometer (MODIS) available for the vegetation period (March, April) for each studied year (2011, 2012, 2016, 2017) at a 250 × 250 m^2^ resolution for our study area.
Since three NDVI scenes were available for each vegetation period, we computed the mean NDVI for each vegetation period.

Please note also that our GAM model clearly reveals a relationship between the interaction of the MODIS NDVI and year and the response variable (adjusted R squared: `r round(summary(gam_1)$r.sq, 2)`) proving that the MODIS data is appropriate for the chosen modeling approach.

7. **Line 105: This is an interesting integrative framework, but it is not clear to me what the source of species in each one plot is after the experiment was establishment. Where they planted as seeds? Is the result of immigration from nearby forest remnants or are artificially included in each plot?**

Thank you very much for the compliment!
The experiment was established on the campus of the University of Piura.
The campus comprises large undisturbed areas mirroring the typical vegetation formation, open xeric shrublands, typical of the region around Piura.
Hence, we only observed how the natural seed bank developed when exposed to artificial water/nutrient input.
Please refer also to the Supplementary Information for a detailed description of the experimental setup.

8. **Line 132: What is Londo-scale?**

Actually, we were referring to the decimal scale proposed by Londo. 
The decimal scale has finer intervals than the Braun-Blanquet scale (Table \@ref(tab:decimal-scale)).
In addition, the decimal scale does not use symbols like `r` and `+` which are hard to handle for software expecting numeric values such as ordination techniques.
We have added following sentence to the description of the vegetation sampling:

>>>Prior to the analysis, we transformed the plant cover values in accordance with the decimal scale by Londo [@londo_decimal_1976].

```{r decimal-scale, echo=FALSE}
tab = tibble(
  "coverage (in %)" = c("<1", "1-3", "3-5", "5-15", "15-25", "25-35", "35-45", "45-55", "55-65", "65-75", "75-85", "85-95", "95-100"),
  "londo_values" = c(0.1, 0.2, 0.4, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10))
cap = "Decimal scale as proposed by Londo (1976)."
kable(tab, "latex", booktabs = TRUE, escape = TRUE, 
      # align = "c",
      # col.names = linebreak(nms, align = "c"),
      caption = cap)
```

9. **Line 138: The modeling approach maybe inappropriate. The authors used a detrended canonical analysis to reduce dimensionality of the community matrix and then used a generalized additive model to compare the DCA axis with environmental data. There are other methods more appropriate to evaluate the effect of climate and other factors on species composition such as canonical correspondence analysis or redundancy analysis. Also, the generalized dissimilarity modeling allows to account for non-linear relationships among predictors and species composition variation.**

Apparently, we needed to rewrite our methods section to communicate more clearly how we have addressed methodologically our aims.
We have now made crystal-clear in the text that we were interested in spatially predicting the floristic gradient represented by the scores of the first DCA axis.
Kindly refer to the completely rewritten section "Modeling and predictive mapping" which now reads as follows:

>>>One aim of this study was to spatio-temporally predict ecosystem diversity (floristic composition) represented by ordination scores along the climatic gradient in the study area.
@muenchow_coupling_2013 have already shown that NDVI is an excellent predictor for vegetation assemblages in the study area.
Therefore, we modeled the scores of the first DCA axis representing the main observed floristic gradient in the study area with the help of a Generalized Additive Model (GAM) as a function of the interaction between NDVI and year:

>>>$$
\begin{aligned}
\text{Scores}_{i}\ \sim\ \mathrm{N}\left( \mu_{i},\ \sigma^{2} \right) \\
\mu_{i} = \alpha + f_{\text{year}_{i}}\left( \text{NDVI}_{i} \right) + \beta_{1}\text{year}_{i}
\end{aligned}
$$

>>>(Formula 1)
>>>
>>>where *i* refers to the *i*th observation, *$\alpha$* is the intercept, *$\beta$*~1~ is the estimated coefficient for year and *f* is a spline smoother for the NDVI:year interaction.
>>>
>>>The sole purpose of the model is the spatial prediction of the ordination scores, not statistical inference or the interpretation of the models' coefficients.
We will use variation partitioning later on to assess how the influence of environmental predictors has changed over the studied years (see also next section). 
Consequently, here we are only interested in a good predictive performance of the model to make sure that it is able to generalize, i.e., that it is able to predict reasonably well unseen data [@james_introduction_2013; @kuhn_applied_2013]. 
Only this can be the basis for a reliable map of the predicted ordination scores.
To make sure this is the case, we used spatial cross-validation.
Overall, cross-validation accounts for over-optimistic predictions on the training set by randomly splitting a data set into partitions used for training and testing [@brenning_spatial_2012;@lovelace_geocomputation_2019a]. 
Spatial cross-validation is an extension that measures a model's ability to spatially predict the response variable in the presence of spatial autocorrelation [@schratz_hyperparameter_2019]. 
We ran a 100-repeated 5-fold spatial cross-validation in which the partitioning is based on *k*-means clustering (*k* = 5; @russ_data_2010).
We also made sure that all annual observations of one plot were jointly placed in either the test or the training data set [@meyer_improving_2018]. 
We used the normalized root mean-square error (NRMSE) in percent as performance measure for the spatial cross-validation.
>>>
>>>$$
NRMSE = \frac{\sqrt{\frac{\sum_{i = 1}^{n}{(e_{i} - y_{i})²}}{n}}}{max(y) - \ min(y)}*100
>>>$$ 
>>>(Formula 2)

>>>where *n* is the number of observations, *e* corresponds to the residuals, and *y* refers to the observed values.

Since spatial prediction was the aim, we were not very much interested in the effect of the predictors (climate, NDVI) but only in a good predictive performance of the model to ensure that it is able to generalize [@james_introduction_2013].
But please note that the variation partioning uses the RDA approach, as suggested, to assess the impact of environmental predictor groups on the floristic composition across time (see also reply to your next comment).

Of course, we are also aware of canonical (constrained) ordination techniques [@legendre_numerical_2012]. 
Constrained ordination techniques model every single species as a function of environmental predictors, and then use the fitted values as input for an ordination method - either a PCA (then called RDA) or a CA (then called CCA).
One shortcoming of these techniques is that they usually explain less of the observed variance.
Another reason why we have not used constrained techniques for the predictive modeling was that we would have to use the same predictors (NDVI) twice - once as input for the CCA/RDA and then again for modeling the scores of the CCA/RDA.

<!--
One shortcoming of constrained techniques is that they exclusively use linear models, which might not be the most appropriate choice for each species encountered along a pronounced environmental gradient [@borcard_numerical_2018].
Secondly, in our case, species were recorded using the Londo scale which is a categorical variable. 
However, a linear model expects a response variable that follows a Gaussian distribution.
That alone would make the usage of a constrained ordination technique at least questionable.
Still, we could use one, and though we would not very much trust the modeled values of a canonical ordination technique for the reasons listed above, we could still use its scores as a response variable instead of the scores of a DCA.
But again, there are at least two reasons why we would refrain from such an approach:

1. Constrained techniques usually explain less of the observed variance of the original input matrix compared to their unconstrained counterparts since they only include in the ordination what the model was able to explain, the rest, i.e., the residuals, are simply ignored, which of course is not the case in the case of unconstrained techniques.
1. We would have to use the same predictors (NDVI and time) for the canonical ordination and the subsequent score modeling, i.e., we would e.g., model the CCA scores which themselves are the fitted values based on the same predictors. 
This smells very much like circular reasoning.
--> 
Finally, the generalized dissimilarity modeling approach is interesting, as it allows to model species turnover (beta diversity), and in fact we will give it a try in a future study.
However, in this study, we were interested in modeling the spatio-temporal change of the vegetation assemblages along a pronounced climatic gradient during four different ENSO years.

10. **Line 156: The variance partitioning approach used in not clear. If it is based on the GAM models employed previously, it is incorrect because they are based on the likelihood of the model (explained deviance) that is similar to R2 in an ordinary least square model, but not the same. In addition, the redundancy analysis approach (RDA) is more frequently used to assess the effect of environment on species composition. Also, the use of similar modeling methods previously employed allow comparability with other studies.**

We are sorry again that obviously we have not presented our methodological approach in sufficient detail.
First, the variation partitioning is not based on the GAM.
Instead, it is based, as suggested in your comment, on an RDA approach.
The variation partitioning computes with the help of RDA how the influence of predictor groups - in our case edaphic and topographic variables - has changed with different levels of water availability [@borcard_numerical_2018].
<!--
Second, you are right that RDA is used to assess the effect of environmental predictors on the species composition.
However, a RDA is not able to quantifiy the effects of different predictor groups - in our case edaphic and topographic variables - on species composition over time.
-->
We now made our reasoning now much clearer in the methods sections.
The updated text in the manuscript reads as follows:

### 2020-08-10 text/ecography/round_2/enso_comp.Rmd
migrate gitlab repo to github

---
output:
  bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: xelatex
    # float is needed for floating figures
    extra_dependencies: ["fontspec", "sectsty", "float"]
    toc: false
    fig_caption: yes
    number_sections: true
bibliography: transect_refs.bib
csl: ecography.csl
---

@@

![Caption for the picture.](/home/jannes/work/acad/science/projects/ecology/latin_america/peru/transect_piura/transect_17/figures/12_map_grid.png)

@@

ENSO is a recurrent climate phenomenon causing anomalies around the globe [@capotondi_understanding_2015].
EN represents the warm and LN the cold phase of this cycle. 
ENSO neutral refers to periods when neither EN nor LN are present. 
LN usually causes the reverse effects of EN. 
The effects depend on the region and include among others excessive rainfalls, droughts as well as temperature, wind and monsoon anomalies. 
<!--
The anomalies show recurrent patterns to a certain degree but are still highly variable especially on a local scale. 
For instance, in 2016 the predicted Super-Niño rainfalls largely failed to materialize along the coasts of Ecuador and Peru.
Just a year later, unexpected, torrential rainfalls hit the same coasts during a supposedly ENSO neutral year with a magnitude comparable to that of the Super-Niño event in 1997/98. Originating from a so-called coastal EN, the 2017 rainfalls came as a complete surprise, and caused the unprepared population unbearable hardship [@fraser_peru_2017; @ramirez_understanding_2017].
-->

Though extreme climate events like EN and LN change and reshape ecosystems drastically [@Kogan.2017], ecological research mostly focusses on the effects of EN.
Nevertheless, the eco-climatic impact of LN is comparable in magnitude with that of EN.
For instance, LN episodes were most likely the main driver for the hiatus in global warming at the beginning of the 21^st^ century [@kosaka_recent_2013].
In any case, ENSO certainly drives a large amount of the interannual variability of woody plant growth across the tropics [@rifai_enso_2018] though the specific response to El Niño may vary from region to region [@phillips_changes_2009].
Overall, the effect of  extreme events is even pronounced in (semi)-arid ecosystems [@holmgren_extreme_2006], which is especially true for ENSO's terrestrial core region in NW Peru and SW Ecuador [@muenchow_woody_2013; @rollenbeck_climatic_2015] -- one of the most variable regions on Earth in terms of climate [@wehrden_global_2012].
This region is also home to the highly tropical dry forest (TDF) ecosystems.
TDFs are characterized by a high rate of endemism and feature an outstanding biodiversity which is almost as high as that of humid tropical forests [@espinosa_what_2011;@pennington_tropical_2018]. 
At the same time, they are the most threatened tropical ecosystem [@miles_global_2006]. 
In Peru, 95% of TDFs have already disappeared [@portillo-quintero_extent_2010]. 
Even so, tropical dry ecosystems are chronically understudied [@escribano-avila_biodiversity_2017;@muenchow_review_2018].

@@

To tackle the outlined issues at least in part, we examined how different ENSO episodes affect the floristic composition along a spatio-temporal gradient in NW Peru. 
For this, we surveyed the floristic composition on permanent plots in four years representing different ENSO episodes (dry (2011) and humid (2012) LN year; moderate EN year (2016); very humid coastal EN year (2017)).
Specifically, we investigated the:

@@

\begin{figure}[H]
\includegraphics[width=25in]{/home/jannes/work/acad/science/projects/ecology/latin_america/peru/transect_piura/transect_17/figures/12_map_grid.png}
\caption{The study area.}
\label{fig:study-area-map}
\end{figure}

@@

The study area is part of the core region where the ENSO phenomenon exerts its greatest terrestrial influence [@richter_monitoring_2005]. 
Generally, EN brings more rain than usual and LN enforces the already dry conditions in the study area.
However, even this pattern is highly variable and can occasionally be reversed, see e.g., the years 2005, 2008 and 2012 in Fig. \@ref(fig:study-area-map)d. Additionally, so-called "local" or "coastal EN events" can take place as observed in 2017. 
Coastal EN events refer to the regional warming of coastal waters in the Gulf of Guayaquil in December as opposed to the Pacific basin-wide EN when warm North Australian waters reach the South American coast.

@@

Therefore, to put the ecological results into context, it is important to keep in mind the varying rainfall patterns as produced by different ENSO episodes across the studied years. 
The observed years differed largely in terms of precipitation.
2011 corresponded to a dry LN year and the whole study area experienced drier conditions than normal.
By contrast, 2012 was a humid LN year, and 1.5 times and 3 times more rain fell in Piura and Chulucanas, i.e., in the middle to eastern part of the study area (Fig. \@ref(fig:study-area-map)).
2016 was a moderate EN year and rainfall was double the median input in Paita and Piura, whereas in Chulucanas the median value was almost reached. 
2017 coincided with a coastal EN event and is the only studied year during which the whole study area experienced exceptionally wet conditions: Paita received almost `r rain_17[rain_17$station == "Paita", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Paita", "precip"] %>% pull %>% round` mm), Piura almost `r rain_17[rain_17$station == "Piura", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Piura", "precip"] %>% pull %>% round` mm) and Chulucanas `r rain_17[rain_17$station == "Chulucanas", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Chulucanas", "precip"] %>% pull %>% round` mm) more rain than on median (Fig. \@ref(fig:study-area-map)d). 
Moreover, it is noteworthy that vegetation development changes drastically in the study area during wet episodes but remains more or less uniform during neutral years [@holmgren_extreme_2006; @muenchow_coupling_2013; @richter_monitoring_2005].

@@

In addition to the plot survey at the landscape scale, we conducted a small-scale irrigation-nutrient experiment to quantify the beneficial effects of the water-nutrient interaction on vegetation development and biomass production.
We established the experiment in the middle of the precipitation gradient, ranging from `r rain_17[rain_17$station == "Paita", "median"] %>% pull %>% round` mm (Paita; annual median precipitation) to `r rain_17[rain_17$station == "Chulucanas", "median"] %>% pull %>% round` mm (Chulucanas; Fig. \@ref(fig:study-area-map)), on the campus of the University of Piura (5°10'S, 80°38'W, see Piura in Fig. \@ref(fig:study-area-map)).
The environmental conditions are typical for the region in terms of climate, topography and soil properties (see @muenchow_soil_2013; @muenchow_woody_2013). 
Inside a sandy area devoid of woody vegetation, we placed 12 experimental plots of 3 × 3 m^2^, which were regularly spaced along a rectangular grid of 1 ha. 
Plots were assigned to one of three irrigation treatments: The first treatment represented a Super EN year (example of 1997/98: 1780 mm), the second represented a moderate EN event (example of 1991/1992: 258 mm). 
The last treatment represented the baseline with no additional water input and corresponded to the year in which the irrigation experiment was conducted (2013) which was a neutral year (54 mm; see also Fig. \@ref(fig:study-area-map)d). 
The irrigation water stemmed from rain-collecting tanks.
In addition to the water treatment, two randomly chosen plots of each irrigation treatment received a fertilization treatment of 200kg/ha granular nitrogen fertilizer (NH~4~NO~3~). 
The experiment was carried out between December 2012 and May 2013.
Plant development was monitored two times per month and included the determination of plant species [@missouribotanicalgarden_tropicos_2018] and corresponding coverage (%).
Please refer to Appendix 2 for a detailed description of the irrigation-nutrient experiment.

@@

All statistical analyses were conducted in the open source software R [@rcoreteam_language_2019] using its packages mgcv [@wood_generalized_2017], RQGIS ([@muenchow_rqgis_2017a], sperrorest [@brenning_spatial_2012] and vegan [@oksanen_vegan_2019]. 
<!-- JM: add link to github repository, zenodo repository, dockerhub -->
To comply with the good practices of reproducibility we will make available the data and the code of this work on an open accessible github repository (https://github.com/) upon acceptance of the manuscript.

@@

```{r pred-map, echo=FALSE, out.width="80%", fig.align="center", fig.cap="Spatial prediction of the floristic composition along the cross-section for different ENSO episodes The urban boundaries of the cities Paita, Piura and Chulucanas are displayed in black. The Piura River and its tributaries are represented by light blue lines with a black contour.", fig.align="centerfig.align=fig.align="}
include_graphics(here("figures/15_pred_map.png"))
```

>>>>

---
output:
  bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: xelatex
    # float is needed for floating figures
    extra_dependencies: ["fontspec", "sectsty", "float"]
    toc: false
    fig_caption: yes
    number_sections: true
header-includes:
  # adjust figure captions
  - \usepackage{caption}
  - \captionsetup{font={footnotesize,bf,singlespacing}}
  # Ecography requires double line spacing and line numbering
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{lineno}
  - \linenumbers

title: "Monitoring and predictive mapping of floristic biodiversity along a climatic gradient in ENSO's terrestrial core region, NW Peru"
bibliography: transect_refs.bib
csl: ecography.csl
# biblio-style: "natbib"
link-citations: true
---

@@

@@

ENSO is a recurrent climate phenomenon causing anomalies around the globe [@capotondi_understanding_2015].
El Niño represents the warm and La Niña the cold phase of this cycle. 
ENSO neutral refers to periods when neither El Niño nor La Niña are present. 
La Niña usually causes the reverse effects of El Niño. 
The effects depend on the region and include among others excessive rainfalls, droughts as well as temperature, wind and monsoon anomalies. 
<!--
The anomalies show recurrent patterns to a certain degree but are still highly variable especially on a local scale. 
For instance, in 2016 the predicted Super-Niño rainfalls largely failed to materialize along the coasts of Ecuador and Peru.
Just a year later, unexpected, torrential rainfalls hit the same coasts during a supposedly ENSO neutral year with a magnitude comparable to that of the Super-Niño event in 1997/98. Originating from a so-called coastal EN, the 2017 rainfalls came as a complete surprise, and caused the unprepared population unbearable hardship [@fraser_peru_2017; @ramirez_understanding_2017].
-->

Though extreme climate events like El Niño and La Niña change and reshape ecosystems drastically [@Kogan.2017], ecological research mostly focusses on the effects of El Niño.
Nevertheless, the eco-climatic impact of La Niña is comparable in magnitude with that of El Niño.
For instance, La Niña episodes were most likely the main driver for the hiatus in global warming at the beginning of the 21^st^ century [@kosaka_recent_2013].
In any case, ENSO certainly drives a large amount of the interannual variability of woody plant growth across the tropics [@rifai_enso_2018] though the specific response to El Niño may vary from region to region [@phillips_changes_2009].
Overall, the effect of  extreme events is even pronounced in (semi)-arid ecosystems [@holmgren_extreme_2006], which is especially true for ENSO's terrestrial core region in NW Peru and SW Ecuador [@muenchow_woody_2013; @rollenbeck_climatic_2015] -- one of the most variable regions on Earth in terms of climate [@wehrden_global_2012].
This region is also home to the highly tropical dry forest (TDF) ecosystems.
TDFs are characterized by a high rate of endemism and feature an outstanding biodiversity which is almost as high as that of humid tropical forests [@espinosa_what_2011;@pennington_tropical_2018]. 
At the same time, they are the most threatened tropical ecosystem [@miles_global_2006]. 
In Peru, 95% of TDFs have already disappeared [@portillo-quintero_extent_2010]. 
Even so, tropical dry ecosystems are chronically understudied [@escribano-avila_biodiversity_2017;@muenchow_review_2018].

@@

To tackle the outlined issues at least in part, we examined how different ENSO episodes affect the floristic composition along a spatio-temporal gradient in NW Peru. 
For this, we surveyed the floristic composition on permanent plots in four years representing different ENSO episodes (dry (2011) and humid (2012) La Niña year; moderate El Niño year (2016); very humid coastal El Niño year (2017)).
Specifically, we investigated the:

@@

```{r study-area-map, echo=FALSE, fig.cap="The study area (Panel A) located in Peru (Panel B) given as a false-color composite representing the change in NDVI across the observed years. The 2011 NDVI serves as the basis for the interannual change (Panel C). Green pixels occur where the interannual difference in NDVI was greatest between 2017 and 2011. Blue color tones indicate that NDVI differences were greatest between 2012 and 2011 (red: between 2016 and 2011). Panel D: Ratio between annual precipitation and median precipitation (median Paita: 25.9mm, median Piura: 71.4mm, median Chulucanas: 272.4mm) measured at the three automated climatic stations (see section \\\\@ref(environmental-variables)). In case the precipitation was smaller than the median, we computed the ratio as: 1 / (precipitation / median(precipitation)) \\* -1. The years in which vegetation sampling took place are printed in bold (x-axis). The ENSO phases are in correspondence with the ONI index of the NOAA Climatic Prediction Center (http://origin.cpc.ncep.noaa.gov/products/analysis\\_monitoring/ensostuff/ONI\\_v5.php). We have intentionally left out the 1997/1998 Super-Niño event since this would have dominated the y-axis."}
# knitr::include_graphics(here("figures/12_map_grid.png"))
suppressPackageStartupMessages(library("raster"))
library("png")
back = as.raster(readPNG(here("figures/12_ternary_background.png")))
front = as.raster(readPNG(here("figures/12_ternary_foreground.png")))
p = readRDS(here("figures/12_map_grid.rds"))
p
```

@@

The study area is part of the core region where the ENSO phenomenon exerts its greatest terrestrial influence [@richter_monitoring_2005]. 
Generally, El Niño brings more rain than usual and La Niña enforces the already dry conditions in the study area.
However, even this pattern is highly variable and can occasionally be reversed, see e.g., the years 2005, 2008 and 2012 in Fig. \@ref(fig:study-area-map)d. Additionally, so-called "local" or "coastal El Niño events" can take place as observed in 2017. 
Coastal El Niño events refer to the regional warming of coastal waters in the Gulf of Guayaquil in December as opposed to the Pacific basin-wide El Niño when warm North Australian waters reach the South American coast.

@@

Therefore, to put the ecological results into context, it is important to keep in mind the varying rainfall patterns as produced by different ENSO episodes across the studied years. 
The observed years differed largely in terms of precipitation.
2011 corresponded to a dry La Niña year and the whole study area experienced drier conditions than normal.
By contrast, 2012 was a humid La Niña year, and 1.5 times and 3 times more rain fell in Piura and Chulucanas, i.e., in the middle to eastern part of the study area (Fig. \@ref(fig:study-area-map)).
2016 was a moderate El Niño year and rainfall was double the median input in Paita and Piura, whereas in Chulucanas the median value was almost reached. 
2017 coincided with a coastal El Niño event and is the only studied year during which the whole study area experienced exceptionally wet conditions: Paita received almost `r rain_17[rain_17$station == "Paita", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Paita", "precip"] %>% pull %>% round` mm), Piura almost `r rain_17[rain_17$station == "Piura", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Piura", "precip"] %>% pull %>% round` mm) and Chulucanas `r rain_17[rain_17$station == "Chulucanas", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Chulucanas", "precip"] %>% pull %>% round` mm) more rain than on median (Fig. \@ref(fig:study-area-map)d). 
Moreover, it is noteworthy that vegetation development changes drastically in the study area during wet episodes but remains more or less uniform during neutral years [@holmgren_extreme_2006; @muenchow_coupling_2013; @richter_monitoring_2005].

@@

In addition to the plot survey at the landscape scale, we conducted a small-scale irrigation-nutrient experiment to quantify the beneficial effects of the water-nutrient interaction on vegetation development and biomass production.
We established the experiment in the middle of the precipitation gradient, ranging from `r rain_17[rain_17$station == "Paita", "median"] %>% pull %>% round` mm (Paita; annual median precipitation) to `r rain_17[rain_17$station == "Chulucanas", "median"] %>% pull %>% round` mm (Chulucanas; Fig. \@ref(fig:study-area-map)), on the campus of the University of Piura (5°10'S, 80°38'W, see Piura in Fig. \@ref(fig:study-area-map)).
The environmental conditions are typical for the region in terms of climate, topography and soil properties (see @muenchow_soil_2013; @muenchow_woody_2013). 
Inside a sandy area devoid of woody vegetation, we placed 12 experimental plots of 3 × 3 m^2^, which were regularly spaced along a rectangular grid of 1 ha. 
Plots were assigned to one of three irrigation treatments: The first treatment represented a Super El Niño year (example of 1997/98: 1780 mm), the second represented a moderate El Niño event (example of 1991/1992: 258 mm). 
The last treatment represented the baseline with no additional water input and corresponded to the year in which the irrigation experiment was conducted (2013) which was a neutral year (54 mm; see also Fig. \@ref(fig:study-area-map)d). 
The irrigation water stemmed from rain-collecting tanks.
In addition to the water treatment, two randomly chosen plots of each irrigation treatment received a fertilization treatment of 200kg/ha granular nitrogen fertilizer (NH~4~NO~3~). 
The experiment was carried out between December 2012 and May 2013.
Plant development was monitored two times per month and included the determination of plant species [@missouribotanicalgarden_tropicos_2018] and corresponding coverage (%).
Please refer to Appendix 2 for a detailed description of the irrigation-nutrient experiment.

@@

All statistical analyses were conducted in the open source software R [@rcoreteam_language_2019] using its packages mgcv [@wood_generalized_2017], RQGIS ([@muenchow_rqgis_2017a], sperrorest [@brenning_spatial_2012] and vegan [@oksanen_vegan_2019].

@@

```{r pred-map, echo=FALSE, out.width="100%", fig.align="center", message=FALSE, fig.cap="Spatial prediction of the floristic composition along the cross-section for different ENSO episodes The urban boundaries of the cities Paita, Piura and Chulucanas are displayed in black. The Piura River and its tributaries are represented by light blue lines with a black contour."}
# include_graphics(here("figures/15_pred_map.png"))
library("sp")
library("sf")
load(here("figures/15_pred_map.rda"))
print(p)
```
